---
title: "preliminary_analysis_Odonata"
author: "Diana Bowler"
date: "24 januar 2018"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

#Analysis of the Odonata data from Saarland

First step: read in the raw data file from the sMon data portal.
Sheet 5 of the workbook contains the raw data of occurrences.
Each row in the data frame is a species observation.

```{r warning=FALSE}

#library(gdata)
load("derived-data/adult_datafile.RData")
df <- adult_datafile

```


#Data frame formatting

Step one: format the date and extract month, dat and year information.

```{r}

library(lubridate)
df$Date <- as.Date(df$Date,file="%Y-%m-%d")
df$month <- month(df$Date)
df$day <- yday(df$Date)
df$Year <- year(df$Date)

```

Step two: remove those with missing date entries

```{r}

df <- subset(df,!is.na(month))
summary(df)

```


#Data filtering

Filter one: focus on the months that were most sampled. The results suggest this is between April and September.

```{r}

table(df$month)

df <- subset(df, month %in% c(5:9))

```

Filter two: begin the time series from the first year with reasonable amount of data (i.e., number of records). 1981 seems like a good place to start. Also, we cut off 2017 incase the database is not up-to-date yet.

```{r}

table(df$Year)
df <- subset(df, Year>1980)
df <- subset(df, Year<2017)

```

#Data availability/Sampling effort checking

We can look at effort by: total number of records and average number of species seen per sampling day ("list length")

```{r}

library(sparta)

dataDiagnostics(taxa= as.character(df$Species), 
                site= as.character(df$MTB_Q), 
                time_period = df$Date)

#both list length and number of records increase over time

```

Summary info per species

```{r}

library(plyr)
(outSummary <- ddply(df,.(Species),summarise,
                     nuRecs=length(unique(Date)),
                     nuYears=length(unique(Year)),
                     nuSites=length(unique(MTB_Q))))

```

Organise data using the sparta wrapper function

```{r, results='asis'}

source('R/sparta_wrapper_functions.R')

#define a visit
df$visit <- paste(df$MTB_Q,df$Date,df$Beobachter,sep="_")

occMatrix <- getOccurrenceMatrix(df)
listlengthDF <- getListLength(df)

#checks
all(row.names(occMatrix)==listlengthDF$surveyList)
listlengthDF[1:10,]

#plots
library(ggplot2)
qplot(nuSpecies,nuRecords,data=listlengthDF)
qplot(nuSpecies,RpS,data=listlengthDF)

```

Compile data for bugs (for a select species  - here )

```{r}

mySpecies<-"Pyrrhosoma nymphula"

bugs.data <- list(nsite = length(unique(listlengthDF$siteIndex)),
                  nyear = length(unique(listlengthDF$yearIndex)),
                  nvisit = nrow(listlengthDF),
                  site = listlengthDF$siteIndex,
                  year = listlengthDF$yearIndex,
                  yday = listlengthDF$day - median(listlengthDF$day),
                  nuSpecies = log(listlengthDF$nuSpecies) - log(median(listlengthDF$nuSpecies)),
                  nuRecs = log(listlengthDF$nuRecords) - log(median(listlengthDF$nuRecords)),
                  nuSS = listlengthDF$samplingSites - median(listlengthDF$samplingSites),# up to 3
                  expertise = log(listlengthDF$expertise)-median(log(listlengthDF$expertise)),
                  RpS = log(listlengthDF$RpS) - median(log(listlengthDF$RpS)),
                  y = as.numeric(occMatrix[,mySpecies]))

listlengthDF$Species <- bugs.data$y
                  
#need to specify initial values
library(reshape2)
zst <- acast(listlengthDF, siteIndex~yearIndex, value.var="Species",fun=max)
zst [is.infinite(zst)] <- 0
inits <- function(){list(z = zst)}

```

Look at the data:

Phenology:

```{r,results='asis'}

library(ggplot2)

#phenology pattern
library(plyr)
monthSums <- ddply(listlengthDF,.(month,Year),summarise,prevalence=sum(Species))
qplot(month,prevalence,data=monthSums)+facet_wrap(~Year,scale="free_y")#quadratic term seems reasonable

#time series trend
yearSums <- ddply(listlengthDF,.(Year,MTB_Q),summarise,total=sum(Species),
                  totRecs=sum(nuRecords),totSpe=sum(nuSpecies))

qplot(Year,total/totRecs,data=yearSums,colour=MTB_Q)
qplot(Year,total/totSpe,data=yearSums,colour=MTB_Q)

```


Run model with different effort terms

(1) with listlength as effort

```{r}

bugs.data$Effort <- bugs.data$nuSpecies

source('R/BUGS_misc_functions.R')

#specify parameters to monitor
params <- c("phenol.p","phenol2.p","effort.p","psi.fs")

#run model
out1 <- jags(bugs.data, inits=inits, params, "R/BUGS_sparta.txt", n.thin=nt,
               n.chains=3, n.burnin=4000,n.iter=20000,parallel=T)

print(out1,2)



```


(2) with records per species as a measure

```{r}

bugs.data$Effort <- bugs.data$RpS

source('R/BUGS_misc_functions.R')

#specify parameters to monitor
params <- c("phenol.p","phenol2.p","effort.p","psi.fs")

#run model
out1 <- jags(bugs.data, inits=inits, params, "R/BUGS_sparta.txt", n.thin=nt,
               n.chains=3, n.burnin=4000,n.iter=20000,parallel=T)

print(out1,2)


```

(3) add in an experise effect

```{r}

bugs.data$Effort <- bugs.data$nuSpecies

source('R/BUGS_misc_functions.R')

#specify parameters to monitor
params <- c("phenol.p","phenol2.p","effort.p","expert.p","psi.fs")

#run model
out1 <- jags(bugs.data, inits=inits, params, "R/BUGS_sparta.txt", n.thin=nt,
               n.chains=3, n.burnin=4000,n.iter=20000,parallel=T)


```

(4) add in number of sampling sites

```{r}

bugs.data$Effort <- bugs.data$nuSpecies

source('R/BUGS_misc_functions.R')

#specify parameters to monitor
params <- c("phenol.p","phenol2.p","effort.p","effort2.p","ss.p","psi.fs")

#run model
out1 <- jags(bugs.data, inits=inits, params, "R/BUGS_sparta.txt", n.thin=nt,
               n.chains=3, n.burnin=4000,n.iter=20000,parallel=T)

```

(5) all detection variables together

```{r}

bugs.data$Effort <- bugs.data$nuSpecies
bugs.data$Effort_v2 <- bugs.data$RpS

source('R/BUGS_misc_functions.R')

#specify parameters to monitor
params <- c("phenol.p","phenol2.p","effort.p","effort_2.p","expert.p","ss.p","psi.fs")

#run model
out1 <- jags(bugs.data, inits=inits, params, "R/BUGS_sparta.txt", n.thin=nt,
               n.chains=3, n.burnin=4000,n.iter=20000,parallel=T)

print(out1,2)

```


#Run models over multiple species

Pull out 20 most common Species

```{r}

outSummary <- arrange(outSummary,desc(nuRecs))
(mySpecies <- outSummary$Species[1:20])
outSummary$nuRecs[1:20]

```

Write general function:

```{r}

fitBugs<-function(mySpecies,effort="nuSpecies",modelfile="R/BUGS_sparta.txt"){
  
#organise data
bugs.data <- list(nsite = length(unique(listlengthDF$siteIndex)),
                  nyear = length(unique(listlengthDF$yearIndex)),
                  nvisit = nrow(listlengthDF),
                  site = listlengthDF$siteIndex,
                  year = listlengthDF$yearIndex,
                  yday = listlengthDF$day - median(listlengthDF$day),
                  nuSpecies = log(listlengthDF$nuSpecies) - log(median(listlengthDF$nuSpecies)),
                  nuRecs = log(listlengthDF$nuRecords) - log(median(listlengthDF$nuRecords)),
                  nuSS = listlengthDF$samplingSites - median(listlengthDF$samplingSites),# up to 3
                  expertise = log(listlengthDF$expertise)-median(log(listlengthDF$expertise)),
                  RpS = log(listlengthDF$RpS) - median(log(listlengthDF$RpS)),
                  y = as.numeric(occMatrix[,mySpecies]))
listlengthDF$Species <- bugs.data$y
                  
#specify initial values
library(reshape2)
zst <- acast(listlengthDF, siteIndex~yearIndex, value.var="Species",fun=max)
zst [is.infinite(zst)] <- 0
inits <- function(){list(z = zst)}

#fit model
bugs.data$Effort <- bugs.data[[effort]]
bugs.data$Effort_v2 <- bugs.data$RpS

source('R/BUGS_misc_functions.R')

#specify parameters to monitor
params <- c("phenol.p","phenol2.p","effort.p","effort_2.p","expert.p","ss.p","psi.fs")

#run model
out1 <- jags(bugs.data, inits=inits, params, modelfile, n.thin=nt,
               n.chains=3, n.burnin=4000,n.iter=20000,parallel=T)

return(out1)
}


```

Run model with effort with all effort terms
```{r}

#run out each species
out <-llply(mySpecies,function(x){
  fitBugs(mySpecies=x)
})

save(out,file="out_nuSpecies_All_adults.RData")

```


Extract info on each effort term
```{r}

load("model-outputs/out_nuSpecies_All_adults.RData")
out2 <- combineModels(out)
out_nuSpecies <- getBUGSfits(out2,param="effort.p") 
out_RpS <- getBUGSfits(out2,param="effort_2.p")
out_Expertise <- getBUGSfits(out2,param="ss.p") 
out_nuSS <- getBUGSfits(out2,param="expert.p")

#combine all

out <-  rbind(out_nuSpecies,out_RpS,out_Expertise,out_nuSS)

#plotting
ggplot(out)+
  geom_crossbar(aes(x=Species,y=mean,ymin=X2.5.,ymax=X97.5.))+
  coord_flip()+
  facet_wrap(~Param,scales="free")+
  geom_hline(yintercept=0,col="red",linetype="dashed")

```

Get annual indices

```{r}

load("model-outputs/out_nuSpecies_All_adults.RData")
out2 <- combineModels(out)
outAnnual <- getBUGSfits(out2,param="psi.fs") 
outAnnual$Year <- outAnnual$ParamNu + 1982

```

Plotting as separate times series

```{r}

library(ggplot2)
ggplot(outAnnual)+
  geom_line(aes(x=Year,y=mean))+
  geom_ribbon(aes(x=Year,ymin=X2.5., ymax=X97.5.),alpha=0.5)+
  facet_wrap(~Species)+
  theme_classic()+
  ylab("Predicted occupancy proportion")

```

Categorise trends

```{r}

load("model-outputs/out_RpS_Expertise_trends.RData")
out2 <- combineModels(out)
outTrends <- getBUGSfits(out2,param="trend.year") 
outTrends$Trend.type <- NA
outTrends$Trend.type[which(outTrends$X2.5.<0 & outTrends$X97.5.>0)]<- "no change"
outTrends$Trend.type[which(outTrends$X2.5.<0 & outTrends$X97.5.<0)]<- "decreasing"
outTrends$Trend.type[which(outTrends$X2.5.>0 & outTrends$X97.5.>0)]<- "increasing"

```

Plotting as separate times series

```{r}

library(ggplot2)
ggplot(outAnnual)+
  geom_line(aes(x=Year,y=mean,colour=Trend.type))+
  geom_ribbon(aes(x=Year,ymin=X2.5., ymax=X97.5.,fill=Trend.type),alpha=0.5)+
  facet_wrap(~Species)+
  theme_classic()+
  ylab("Predicted occupancy proportion")

```

Plotting as trend type

```{r}

outAnnual$Trend.type <- factor(outAnnual$Trend.type,levels=c("decreasing",
                                                             "no change",
                                                             "increasing"))
library(ggplot2)
ggplot(outAnnual)+
  geom_line(aes(x=Year,y=mean,group=Species))+
  facet_wrap(~Trend.type)+
  theme_classic()+
  theme(legend.position="none")+
  ylab("Predicted occupancy proportion")

```

Model checking

```{r}

print(out1,3)
traceplot(out1)

```

