---
title: "formatting_Saarland"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}

datafile <- read.csv("raw-data/Saarland/Trockur_Saarl_2018_1.csv",as.is=T)

source('C:/Users/db40fysa/Nextcloud/sMon-Analyses/Git/sMon-insects/R/sparta_wrapper_functions.R')

```

Format data info

```{r}

library(lubridate)
datafile$Date <- as.Date(datafile$datum,format="%d.%m.%Y")
datafile$Month <- month(datafile$Date)
datafile$Year <- year(datafile$Date)
datafile$yday <- yday(datafile$Date)

```


Check species names against the standard list

```{r}
specieslist <- read.delim("specieslist_odonata.txt")

unique(datafile$Species[!datafile$Species%in%specieslist$Species])

#remover genus only
datafile <- subset(datafile,grepl(" ",Species))
#correct species names
datafile$Species[which(datafile$Species=="Lestes virens vestalis")]<-"Lestes virens"
datafile$Species[which(datafile$Species=="Lestes viridis")]<-"Chalcolestes viridis"

unique(datafile$Species[!datafile$Species%in%specieslist$Species])
```

Sort data columns

```{r}

#Kopula column
datafile$Kopula <- gsub(">","",datafile$Kopula)
datafile$Kopula <- gsub("mind ","",datafile$Kopula)
datafile$Kopula <- as.numeric(as.character(datafile$Kopula))
datafile$Kopula[is.na(datafile$Kopula)] <- 0

#Larven
datafile$Larven <- gsub("jag","",datafile$Larven)
datafile$Larven <- as.numeric(as.character(datafile$Larven))
datafile$Larven[is.na(datafile$Larven)] <- 0

#exuvien
datafile$Exuvien[which(datafile$Exuvien=="21-50")] <- 21
datafile$Exuvien <- gsub(">","",datafile$Exuvien)
datafile$Exuvien <- as.numeric(as.character(datafile$Larven))
datafile$Exuvien[is.na(datafile$Exuvien)] <- 0

#oviposition
datafile$Eiablage <- as.numeric(as.character(datafile$Eiablage))
datafile$Eiablage[is.na(datafile$Eiablage)] <- 0

#juvenile
datafile$Juvenile[which(datafile$Juvenile=="50-100")] <- 50
datafile$Juvenile <- as.numeric(as.character(datafile$Juvenile))
datafile$Juvenile[is.na(datafile$Juvenile)] <- 0

#adulte
datafile$Adulte <- sapply(datafile$Adulte,function(x)getFirstRange(x))
datafile$Adulte <- gsub(">","",datafile$Adulte)
datafile$Adulte <- gsub("mind. ","",datafile$Adulte)
datafile$Adulte <- as.numeric(as.character(datafile$Adulte))
datafile$Adulte[is.na(datafile$Adulte)] <- 0

```

Identifying adult and juvenile observation

```{r}

#Jagdflug#all good
#Jagend# all good
#Ruhe# all good
datafile$Adult <- apply(datafile[,c("Adulte","Jagdflug","Jagend","Ruhe","Eiablage","Kopula")],1,sum,na.rm=T)

#create a total juvenile column
datafile$Juv <- datafile$Larven + datafile$Exuvien + datafile$Juvenile

```

Restrict to adult sampling

```{r}

datafile <- subset(datafile, Adult>0)

#check for extremes
hist(datafile$Adult)
quantile(datafile$Adult,0.99)

#block upper values
datafile$Adult[datafile$Adult > quantile(datafile$Adult,0.99)] <- quantile(datafile$Adult,0.99)
hist(datafile$Adult)

```

Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(Fundort))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))


library(GGally)
ggpairs(surveys[,5:9])


```

Characterize the total richness found in each grid:

```{r}
gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 49!!!
#1 is unlikely...

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

#add to data frame
datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

summary(datafile$Richness)

```


Get information on number of sampling locations on a given date

```{r}

#do we just have MTB centroids
samplingSites <- ddply(datafile,.(MTB_Q),summarise,
                  nuCoords=length(unique(interaction(lon,lat))),
                  nuLocations=length(unique(Fundort)))
summary(samplingSites$nuLocations)
summary(samplingSites$nuCoords)
#we seem to have multiple

#per visit
samplingSites <- ddply(datafile,. (Date,MTB_Q),summarise,
                  nuCoords=length(unique(interaction(lon,lat))),
                  nuLocations=length(unique(Fundort)))

summary(samplingSites$nuLocations)
summary(samplingSites$nuCoords)
#usually just one or two

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))
#there is an "unbekannt""

datafile$Beobachter[which(datafile$Beobachter=="Uwe Lingenfelder-SB")]<-"Uwe Lingenfelder"
datafile$Beobachter[which(datafile$Beobachter=="Uwe Lingenfelder/SB")]<-"Uwe Lingenfelder"

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)
#1 to 46!!!

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

allObservers <- as.character(sort(unique(datafile$Beobachter)))

allObservers <- removeGerman(allObservers)
allObservers <- formatObservers(allObservers)

#get double counting of surnames
observerSurnames <- sapply(allObservers,getSurnames)
checkDuplicates(observerSurnames)


#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise<-ldply(allObservers,function(x){
  Beobachter <- as.character(datafile$Beobachter)
  Beobachter <- removeGerman(Beobachter)
  Level = sum(grepl(x,Beobachter))
  data.frame(Beobachter=x,Level)
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

#B. is probably Benrd Trockur -but in any case it is the top ranked observer

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <-
  sapply(as.character(datafile$Beobachter),function(x){
    myobs <- removeGerman(x)
    myobs <- formatObservers(myobs)
    max(expertise$Level[expertise$Beobachter %in% myobs])#taken maximum level when there is a group of observers
    })

summary(datafile$Expertise)

```

Get rid of unknown observers (if there are any)

```{r}

datafile <- subset(datafile, Beobachter!="unbekannt")

```

Save juvenile and adult data separately

```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","Fundort","Adult")]
names(adult_datafile)[which(names(adult_datafile)=="Adult")] <- "Anzahl_min"
save(adult_datafile,file="derived-data/adult_datafile_Sa.RData")
saveRDS(adult_datafile,file="derived-data/adult_datafile_Sa.rds")

```



