---
title: "formatting_Saarland"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}

datafile <- read.csv("raw-data/SH/Libellen_Schleswig-Holstein.csv",as.is=T)

source('C:/Users/db40fysa/Nextcloud/sMon-Analyses/Git/sMon-insects/R/sparta_wrapper_functions.R')

```

Format data info

```{r}

library(lubridate)
datafile$Date <- as.Date(datafile$BEO_DATUM,format="%m/%d/%Y")
datafile$Month <- month(datafile$Date)
datafile$Year <- year(datafile$Date)
datafile$yday <- yday(datafile$Date)
datafile$Species <- datafile$species
datafile$Beobachter <- datafile$ERFNAME
datafile$Fundort <- datafile$Name.1

```


Check species names against the standard list

```{r}
specieslist <- read.delim("specieslist_odonata.txt")

unique(datafile$Species[!datafile$Species%in%specieslist$Species])

#fix names
datafile$Species[which(datafile$Species=="Hemianax ephippiger")]<-"Anax ephippiger"
datafile$Species[which(datafile$Species=="Lestes viridis")]<-"Chalcolestes viridis"
datafile$Species[which(datafile$Species=="Aeshna isosceles")]<-"Aeshna isoceles"
datafile$Species[which(datafile$Species=="Enallagma Charpentier")]<-"Enallagma cyathigerum"
datafile<- subset(datafile, Species!="PF Sympetrum")

unique(datafile$Species[!datafile$Species%in%specieslist$Species])

```

Check out the zeros

```{r}

summary(datafile$ANZAHL)

outP <- subset(datafile,ANZAHL>0)
datesP <- unique(outP$Date)

outA<-subset(datafile,ANZAHL==0)
datesA <- unique(outA$Date)

#on all absent dates, was there at least one presence
out <- datesA[!datesA %in% datesP]
#these are mostly early records

length(out)#540
length(datesP)#4096

#zeros are probably presence only data without a count
#count is unknown, presence is not
datafile$ANZAHL[datafile$ANZAHL==0]<-NA

```

Restrict to juvenile sampling

```{r}

datafile <- subset(datafile, VERH %in% 
                     c("I","PA","EI"))#imagos, pairs copulating, laying eggs

#check for extremes
hist(datafile$ANZAHL)
quantile(datafile$ANZAHL,0.99,na.rm=T)

#block upper values
datafile$ANZAHL[datafile$ANZAHL > quantile(datafile$ANZAHL,0.99,na.rm=T)] <- quantile(datafile$ANZAHL,0.99,na.rm=T)
hist(datafile$ANZAHL)

```


Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(Fundort))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))


library(GGally)
ggpairs(surveys[,5:9])


```

Characterize the total richness found in each grid:

```{r}
gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 47!!!

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

#add to data frame
datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

summary(datafile$Richness)


```


Get information on number of sampling locations on a given date

```{r}

samplingSites <- ddply(datafile,. (Date,MTB_Q),summarise,
nuLocations=length(unique(interaction(lon,lat))))

summary(samplingSites$nuLocations)

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))
#34

table(datafile$Beobachter)
#most collected by Arne Drews

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

allObservers <- as.character(sort(unique(datafile$Beobachter)))

allObservers <- removeGerman(allObservers)
allObservers <- formatObservers(allObservers)

#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise<-ldply(allObservers,function(x){
  Beobachter <- as.character(datafile$Beobachter)
  Beobachter <- removeGerman(Beobachter)
  Level = sum(grepl(x,Beobachter))
  data.frame(Beobachter=x,Level)
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

#observer called "??????", give token expertise of 1
expertise$Level[expertise$Beobachter=="??????"]<-1

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <-
  sapply(as.character(datafile$Beobachter),function(x){
    myobs <- removeGerman(x)
    myobs <- formatObservers(myobs)
    max(expertise$Level[expertise$Beobachter %in% myobs])#take maximum level when there is a group of observers
    })

datafile$Expertise[is.infinite(datafile$Expertise)]<-1

summary(datafile$Expertise)


```


Save data

```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","ANZAHL")]
names(adult_datafile)[which(names(adult_datafile)=="ANZAHL")] <- "Anzahl_min"
save(adult_datafile,file="derived-data/adult_datafile_SH.RData")
saveRDS(adult_datafile,file="derived-data/adult_datafile_SH.rds")

```



