---
title: "formatting_Bavaria"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}
library(rgdal)
datafile <- readOGR(dsn="C:/Users/db40fysa/Nextcloud/sMon-Analyses/Git/sMon-insects/raw-data/Saschen-Anhalt/Libellen_DatenbankLAU_Gesamt_20190208",
                    layer="Libellen_DatenbankLAU_Gesamt_20190208")

#"+proj=utm +zone=32 +ellps=GRS80 +units=m +no_defs"
```

Convert it into a normal dataframe

```{r}
datafile <- data.frame(datafile@data,datafile@coords)

```


Check species names against the standard list

```{r}
specieslist <- read.delim("specieslist_odonata.txt")
datafile$Species <- datafile$Art_wiss
unique(datafile$Species[!datafile$Species%in%specieslist$Species])
datafile <- subset(datafile, !Species%in% c("- Negativnachweis -",
                                            "Gomphus spec.",
                                            "Sympetrum"))
```
Fix MTBQ data

```{r}

datafile$MTB_Q <- paste0(as.character(datafile$MTB),as.character(datafile$MTBQ))

```

Format data info

```{r}
library(lubridate)
datafile$Month <- datafile$Monat
datafile$Year <- datafile$Jahr
datafile$Date <-sapply(datafile$Datum,function(x)strsplit(as.character(x)," ")[[1]][1])
datafile$Date <- as.Date(datafile$Date,format="%d.%m.%Y")
datafile$yday <- yday(datafile$Date)

```

Standardize other headings

```{r}

datafile$Fundort <-paste(datafile$coords.x1,datafile$coords.x2)
datafile$Beobachter <- as.character(datafile$ID.Observ) 

#change coordinates to long lat
#EPSG:31468
library(sp)
library(rgdal)
mycoords<- datafile[,c("coords.x1","coords.x1")]
coordinates(mycoords)<-c("coords.x1","coords.x1")
proj4string(mycoords)<-CRS("+proj=utm +zone=32 +ellps=GRS80 +units=m +no_defs")
mycoords <- spTransform(mycoords,CRS("+proj=longlat +datum=WGS84"))
datafile$lon <- mycoords@coords[,1]
datafile$lat <- mycoords@coords[,2]

```


Sort stage info
```{r}

#stage info here
unique(datafile$Einheit)
# [1] Alttier/Imago                      Exuvien                           
# [3] <NA>                               Individuum (unbest. Altersklassen)
# [5] Emergenz                           Larve                             
# [7] Jungtier                           Eiablage                          
# [9] Kopula                             Kokon                             
#[11] keine Angaben 

table(datafile$Einheit)

#restrict to adults
datafile <- subset(datafile, Einheit %in% c("Alttier/Imago","Kopula",
                                            "Individuum (unbest. Altersklassen)"))

```


Sort data columns

```{r}
#there is no abundance data

datafile$Anzahl_min<-NA

```

Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(Fundort))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))


library(GGally)
ggpairs(surveys[,5:9])


```

Characterize the total richness found in each grid:

```{r}


gridRichness <- ddply(datafile,.(MTB_Q,Month),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))


qplot(Month, nuSpecies,data=gridRichness)

gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 39

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

```


Get information on number of sampling locations on a given date

```{r}

samplingSites <- ddply(datafile,. (Date,MTB_Q),summarise,
nuLocations=length(unique(interaction(lon,lat))))

summary(samplingSites$nuLocations)
#between 1 and 2, usually 1

#add to the data frame
datafile$samplingSites <- samplingSites$nuLocations[match(interaction(datafile$Date,datafile$MTB_Q),
                                interaction(samplingSites$Date,samplingSites$MTB_Q))]

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)
#1 to 44

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

source('R/sparta_wrapper_functions.R')

#extract duplicates
library(gdata)

allObservers <- as.character(sort(unique(datafile$Beobachter)))
allObservers <- removeGerman(allObservers)

#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise<-ldply(allObservers,function(x){
  Beobachter <- as.character(datafile$Beobachter)
  Beobachter <- removeGerman(Beobachter)
  Level = sum(Beobachter==x)
  data.frame(Beobachter=x,Level)
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <-
  sapply(as.character(datafile$Beobachter),function(x){
    myobs <- removeGerman(x)
    max(expertise$Level[expertise$Beobachter %in% myobs])#take maximum level when there is a group of observers
    })

datafile$Expertise[is.infinite(datafile$Expertise)]<-1

summary(datafile$Expertise)

```


Save data

```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","Anzahl_min")]
save(adult_datafile,file="derived-data/adult_datafile_SAnhalt.RData")
saveRDS(adult_datafile,file="derived-data/adult_datafile_SAnhalt.rds")

```

