---
title: "formatting_Bavaria"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}

datafile <- read.csv("raw-data/NRW/rohdaten_aklnrw_an_idiv_181116/data-1542363531429.csv",as.is=T)
nrow(datafile)#177205

```

Add species names

```{r}
speciesDF <- read.csv("raw-data/NRW/rohdaten_aklnrw_an_idiv_181116/reflibellenarten.csv",as.is=T)
datafile$Species <- speciesDF$wiss_name[match(datafile$art,speciesDF$art_id)]

```

Check species names against the standard list

```{r}
specieslist <- read.delim("specieslist_odonata.txt")

unique(datafile$Species[!datafile$Species%in%specieslist$Species])
datafile$Species[which(datafile$Species=="Lestes viridis")]<-"Chalcolestes viridis"

unique(datafile$Species[!datafile$Species%in%specieslist$Species])
```

Check out geographic coords

```{r}
geoDF <- read.csv("raw-data/NRW/rohdaten_aklnrw_an_idiv_181116/reffundortart.csv",as.is=T)

datafile$fundortart <- geoDF$erlaeuterung[match(datafile$fundortart,geoDF$fundortartid)]
table(datafile$fundortart)

#remove those without exact coords??
datafile <- subset(datafile,fundortart!="Fundort ohne (genaue) Koordinatenangaben")

```

Add MTB_Q info

```{r}
library(rgdal)

#get MTB Q
MTBshapefile <- readOGR(dsn="C:/Users/db40fysa/Nextcloud/sMon-Analyses/MTB_Q Informations/MTBQ_shapefile",layer="MTBQ_25833")
#crs specified as +proj=utm +zone=32 +ellps=GRS80 +units=m +no_defs
MTBshapefile@data$Q <- NA
MTBshapefile@data$Q[which(MTBshapefile@data$Quadrant=="NW")]<-1
MTBshapefile@data$Q[which(MTBshapefile@data$Quadrant=="NO")]<-2
MTBshapefile@data$Q[which(MTBshapefile@data$Quadrant=="SW")]<-3
MTBshapefile@data$Q[which(MTBshapefile@data$Quadrant=="SO")]<-4
MTBshapefile@data$MTBQ <- paste0(as.character(MTBshapefile@data$Value),
                                 as.character(MTBshapefile@data$Q))

#overlap coordinates with the grid reference system
mycoords <- datafile[,c("r_wert","h_wert")]
coordinates(mycoords) <- c("r_wert","h_wert")
proj4string(mycoords)<-CRS("+proj=tmerc +lat_0=0 +lon_0=6 +k=1 +x_0=2500000 +y_0=0 +ellps=bessel +datum=potsdam +units=m +no_defs")
mycoords <- spTransform(mycoords,CRS(proj4string(MTBshapefile)))
#proj4string(mycoords)<-CRS("+proj=utm +zone=32 +ellps=GRS80 +units=m +no_defs")
head(mycoords@coords)

#overlay
plot(MTBshapefile)
plot(mycoords,add=T,colour="red")

#looks good!
datafile$MTB <- as.numeric(over(mycoords,MTBshapefile)[,5])
mycoords <- spTransform(mycoords,CRS("+proj=longlat +datum=WGS84"))
datafile$lon <- mycoords@coords[,1]
datafile$lat <- mycoords@coords[,2]
datafile <- subset(datafile, !is.na(MTB))
```

Sort out the dates

```{r}
library(lubridate)
datafile$Date <- as.Date(datafile$datum) 
datafile$Month <- month(datafile$Date)
datafile$Year <- year(datafile$Date)
datafile$yday <- yday(datafile$Date)
summary(datafile$yday)

nrow(subset(datafile,is.na(yday)))
nrow(subset(datafile,!is.na(yday)))

datafile <- subset(datafile, !is.na(yday))

```

Standardize other headings

```{r}

datafile$MTB_Q <- datafile$MTB
datafile$Fundort <-paste(datafile$r_wert,datafile$h_wert)
datafile$Beobachter <- datafile$quelleid

```


Sort stage info and restrict to adult

```{r}

datafile <- subset(datafile,imag_kl>0)

```


Sort data columns

```{r}

hist(datafile$imag_kl)
summary(datafile$imag_kl)

```

Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(Fundort))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))


library(GGally)
ggpairs(surveys[,5:9])


```

Characterize the total richness found in each grid:

```{r}


gridRichness <- ddply(datafile,.(MTB_Q,Month),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))


qplot(Month, nuSpecies,data=gridRichness)

gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 48

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

```


Get information on number of sampling locations on a given date

```{r}

samplingSites <- ddply(datafile,. (Date,MTB_Q),summarise,
nuLocations=length(unique(interaction(lon,lat))))

summary(samplingSites$nuLocations)
#between 1 and 88

#add to the data frame
datafile$samplingSites <- samplingSites$nuLocations[match(interaction(datafile$Date,datafile$MTB_Q),
                                interaction(samplingSites$Date,samplingSites$MTB_Q))]

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)
#1 to 70

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

source('R/sparta_wrapper_functions.R')

#extract duplicates
library(gdata)

allObservers <- as.character(sort(unique(datafile$Beobachter)))
allObservers <- removeGerman(allObservers)

#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise<-ldply(allObservers,function(x){
  Beobachter <- as.character(datafile$Beobachter)
  Beobachter <- removeGerman(Beobachter)
  Level = sum(Beobachter==x)
  data.frame(Beobachter=x,Level)
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <-
  sapply(as.character(datafile$Beobachter),function(x){
    myobs <- removeGerman(x)
    max(expertise$Level[expertise$Beobachter %in% myobs])#take maximum level when there is a group of observers
    })

datafile$Expertise[is.infinite(datafile$Expertise)]<-1

summary(datafile$Expertise)

```


Save data

```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","imag_kl")]
names(adult_datafile)[which(names(adult_datafile)=="imag_kl")] <- "Anzahl_min"
save(adult_datafile,file="derived-data/adult_datafile_NRW.RData")
saveRDS(adult_datafile,file="derived-data/adult_datafile_NRW.rds")

```

