---
title: "formatting_Bavaria"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}

datafile <- read.csv("raw-data/BB/daten_BB_mtb.csv",as.is=T,header=F)
nrow(datafile)#122945
head(datafile)

```

Add in headers

```{r}

#strangely the headings are in the middle of the data frame rows
names(datafile) <- datafile[6301,]
datafile <- datafile[-6301,]

```

Check species names against the standard list

```{r}
specieslist <- read.delim("specieslist_odonata.txt")
datafile$Species <- paste(datafile$GATTUNG,datafile$ART)
unique(datafile$Species[!datafile$Species%in%specieslist$Species])

#fix differences
datafile$Species[which(datafile$Species=="Lestes viridis")]<-"Chalcolestes viridis"
datafile$Species[which(datafile$Species=="Aeshna subarctica elisabeth")]<-"Aeshna subarctica"
datafile$Species[which(datafile$Species=="Aeshna subarctica elisabethae")]<-"Aeshna subarctica"
datafile$Species[which(datafile$Species=="Lestes virens vestalis")]<-"Lestes virens"
#check all fixed
unique(datafile$Species[!datafile$Species%in%specieslist$Species])

```

Format data info

```{r}

library(lubridate)
datafile$Date <- as.Date(datafile$BASIS_datum1,format="%m/%d/%Y")
datafile$Month <- month(datafile$Date)
datafile$Year <- year(datafile$Date)
datafile$yday <- yday(datafile$Date)

#how many have the 1st Jan as a date
nrow(subset(datafile,yday==1))
datafile <- subset(datafile,yday!=1)

```

How does number of sampling years at each square change over time?

```{r, results="asis"}

library(plyr)
siteFreq <- ddply(datafile,.(mtb,Year),summarise,
                  nuVisits=length(unique(yday)))

library(ggplot2)
qplot(Year,nuVisits,data=subset(siteFreq,Year>1980),colour=mtb)+
  theme(legend.position="none")+
  scale_y_log10()

```

Standardize other headings

```{r}

datafile$MTB <- sapply(datafile$mtb,function(x)strsplit(x,"\\.")[[1]][1])
datafile$Q <- sapply(datafile$mtb,function(x)strsplit(x,"\\.")[[1]][2])
datafile$Q <- sapply(datafile$Q,function(x)substr(x,1,1))
datafile$MTB_Q <- paste0(datafile$MTB,datafile$Q)
datafile$Fundort <-datafile$BASIS_ort
datafile$Beobachter <- datafile$BASIS_beobachter 

#how many MTB_Q's are actually MTBs..
sum(!grepl("\\.",datafile$mtb))#most do
sum(grepl("\\.",datafile$mtb))

#remove those without a .
datafile <- datafile[grepl("\\.",datafile$mtb),]
datafile <- subset(datafile, !is.na(Q))

```

Get lon/lat of MTBs

```{r}

library(rgdal)
library(sp)
mtbqs <- readOGR(dsn="C:/Users/db40fysa/Nextcloud/sMon-Analyses/MTB_Q Informations/MTBQ_shapefile",layer="MTBQ_25833")
mtbqs <- spTransform(mtbqs,
CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
mtbqsDF <- data.frame(mtbqs@data,coordinates(mtbqs))

mtbqsDF$Q <- NA
mtbqsDF$Q[which(mtbqsDF$Quadrant=="NW")]<-1
mtbqsDF$Q[which(mtbqsDF$Quadrant=="NO")]<-2
mtbqsDF$Q[which(mtbqsDF$Quadrant=="SW")]<-3
mtbqsDF$Q[which(mtbqsDF$Quadrant=="SO")]<-4
mtbqsDF$MTB_Q <- paste0(as.character(mtbqsDF$Value),
                                 as.character(mtbqsDF$Q))

datafile$lat <- mtbqsDF$X1[match(datafile$MTB_Q,mtbqsDF$MTB_Q)]
datafile$lon <- mtbqsDF$X2[match(datafile$MTB_Q,mtbqsDF$MTB_Q)]
summary(datafile$lat)
summary(datafile$lon)

```


Sort stage info
```{r}

#stage info here
unique(datafile$BASIS_stadium)
#[1] "Imago"                      "Exuvie"                     "Larve"                     
#[4] "juvenil, frisch geschlüpft" "juvenil, frischgeschlüpft" 

table(datafile$BASIS_stadium)

#restrict to adults
datafile <- subset(datafile,  BASIS_stadium%in% c("Imago"))


```


Sort data columns
its only presence data

```{r}

datafile$Anzahl_min <- NA

```

Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(Fundort))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))

library(GGally)
ggpairs(surveys[,5:9])


```

Characterize the total richness found in each grid:

```{r}


gridRichness <- ddply(datafile,.(MTB_Q,Month),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))


qplot(Month, nuSpecies,data=gridRichness)

gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 54

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

```


Get information on number of sampling locations on a given date

```{r}

samplingSites <- ddply(datafile,. (Date,MTB_Q),summarise,
nuLocations=length(unique(Fundort)))

summary(samplingSites$nuLocations)
#always 1

#add to the data frame
datafile$samplingSites <- samplingSites$nuLocations[match(interaction(datafile$Date,datafile$MTB_Q),
                                interaction(samplingSites$Date,samplingSites$MTB_Q))]

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}


source('R/sparta_wrapper_functions.R')

#extract duplicates
library(gdata)

allObservers <- as.character(sort(unique(datafile$Beobachter)))
allObservers <- removeGerman(allObservers)
allObservers <- formatObservers(allObservers)

#get double counting of surnames
#observerSurnames <- sapply(allObservers,getSurnames)
#checkDuplicates(observerSurnames)

#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise<-ldply(allObservers,function(x){
  Beobachter <- as.character(datafile$Beobachter)
  Beobachter <- removeGerman(Beobachter)
  Beobachter <- formatObservers(Beobachter)
  Level = sum(Beobachter==x)
  data.frame(Beobachter=x,Level)
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

expertise$Level[expertise$Beobachter=="ANONYMUS"]<-1
expertise$Level[expertise$Beobachter=="keine Angabe"]<-1

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <-
  sapply(as.character(datafile$Beobachter),function(x){
    myobs <- removeGerman(x)
    max(expertise$Level[expertise$Beobachter %in% myobs])#take maximum level when there is a group of observers
    })

datafile$Expertise[is.infinite(datafile$Expertise)]<-1

summary(datafile$Expertise)

```


Save data

```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","Anzahl_min")]
save(adult_datafile,file="derived-data/adult_datafile_BB.RData")
saveRDS(adult_datafile,file="derived-data/adult_datafile_BB.rds")

```

