---
title: "formatting_Saarland"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}

library(rgdal)
datafile <- readOGR(dsn="C:/Users/db40fysa/Nextcloud/sMon-Analyses/Git/sMon-insects/raw-data/RLP/direct_download/export_artdatenpool",
                     layer="export_artdatenpool")
#+proj=utm +zone=32 +ellps=GRS80 +units=m +no_defs

source('C:/Users/db40fysa/Nextcloud/sMon-Analyses/Git/sMon-insects/R/sparta_wrapper_functions.R')

```

Convert shapefile into data frame

```{r}

datafile <- data.frame(datafile@data,coordinates(datafile))

```


Format data info

```{r}

library(lubridate)
datafile$Date <- as.Date(datafile$k_datum_a,format="%Y/%m/%d")
datafile$Month <- month(datafile$Date)
datafile$Year <- year(datafile$Date)
datafile$yday <- yday(datafile$Date)
datafile$Species <- as.character(datafile$art_wiss)
datafile$Beobachter <- datafile$auftragneh
datafile$Fundort <- interaction(datafile$X1,datafile$X2)
datafile$MTB_Q <- datafile$tk_qu

```


Check species names against the standard list

```{r}
specieslist <- read.delim("specieslist_odonata.txt")

unique(datafile$Species[!datafile$Species%in%specieslist$Species])

#fix names
datafile$Species[which(datafile$Species=="Lestes viridis")]<-"Chalcolestes viridis"
unique(datafile$Species[!datafile$Species%in%specieslist$Species])

```

Its only presence data - add the NAs for abundance

```{r}

datafile$Anzahl_min <- NA

```

Add lon/lat info

```{r}

library(sp)
library(rgdal)
mycoords<- datafile[,c("X1","X2")]
coordinates(mycoords)<-c("X1","X2")
proj4string(mycoords)<-CRS("+proj=utm +zone=32 +ellps=GRS80 +units=m +no_defs")
mycoords <- spTransform(mycoords,CRS("+proj=longlat +datum=WGS84"))
datafile$lon <- mycoords@coords[,1]
datafile$lat <- mycoords@coords[,2]


```



Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(Fundort))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))


library(GGally)
ggpairs(surveys[,5:9])


```

Characterize the total richness found in each grid:

```{r}
gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 47!!!

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

#add to data frame
datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

summary(datafile$Richness)


```


Get information on number of sampling locations on a given date

```{r}

samplingSites <- ddply(datafile,. (Date,MTB_Q),summarise,
nuLocations=length(unique(interaction(lon,lat))))

summary(samplingSites$nuLocations)
#only ever 1 location

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))
#49

#fix mistakes
datafile$Beobachter[which(datafile$Beobachter=="Brown, Gary (Denz, Olaf)")]<-"Brown, Gary"
datafile$Beobachter[which(datafile$Beobachter=="Brown, Gary (Knoblauch, Josef)")]<-"Brown, Gary"
datafile$Beobachter[which(datafile$Beobachter=="Brown, Gary (Steger, Thomas)")]<-"Brown, Gary"
datafile$Beobachter[which(datafile$Beobachter=="L.U.P.O. GmbH - Ott, Jürgen")]<-"L.U.P.O.GmbH - Dr. Ott, Jürgen"
datafile$Beobachter[which(datafile$Beobachter=="Weber, Torsten (Denz, Olaf)")]<-"Weber, Torsten"

table(datafile$Beobachter)
#most collected by KoNat

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

allObservers <- as.character(sort(unique(datafile$Beobachter)))

allObservers <- removeGerman(allObservers)
allObservers <- formatObservers(allObservers)

#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise<-ldply(allObservers,function(x){
  Beobachter <- as.character(datafile$Beobachter)
  Beobachter <- removeGerman(Beobachter)
  Level = sum(grepl(x,Beobachter))
  data.frame(Beobachter=x,Level)
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))


```

Add information on expertise to the datafile

```{r}

datafile$Expertise <-
  sapply(as.character(datafile$Beobachter),function(x){
    myobs <- removeGerman(x)
    myobs <- formatObservers(myobs)
    max(expertise$Level[expertise$Beobachter %in% myobs])#take maximum level when there is a group of observers
    })

datafile$Expertise[is.infinite(datafile$Expertise)]<-1

summary(datafile$Expertise)


```


Save data

```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","Anzahl_min")]

save(adult_datafile,file="derived-data/adult_datafile_RLP.RData")
saveRDS(adult_datafile,file="derived-data/adult_datafile_RLP.rds")

```



