---
title: "formatting_Saarland"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}

datafile <- read.csv("raw-data/ASK_BY_Libellen.csv",as.is=T)
nrow(datafile)#203829

```

How much data has dates?

```{r}

outM <- subset(datafile, mon!=0)
nrow(outM)#176249

#and without day too?
outM <- subset(datafile, mon!=0 & tag!=0)
nrow(outM)#168555

```

What is the difference between von_jahr and bis_jahr

```{r}

outM$diff_jahr <- outM$bis_jahr - outM$von_jahr
table(outM$diff_jahr)

```

How does number of sampling years at each square change over time?

```{r}

library(plyr)
siteFreq <- ddply(outM,.(quadrant,jahr),summarise,
                  nuVisits=length(unique(interaction(tag,mon))))

library(ggplot2)
qplot(jahr,nuVisits,data=subset(siteFreq,jahr>1980),colour=quadrant)+
  theme(legend.position="none")+
  scale_y_log10()

```

Format data info

```{r}

datafile<- outM

library(lubridate)
datafile$Month <- datafile$mon
datafile$Year <- datafile$jahr
datafile$Date <- as.Date(paste(datafile$jahr,datafile$mon,datafile$tag,sep="-"))
datafile$yday <- yday(datafile$Date)

```

Sort data columns

```{r}

hist(datafile$zahl)
summary(datafile$zahl)
#get rid of zeros??
datafile <- subset(datafile,zahl!=0)

#bound extreme values
upperQ <- quantile(datafile$zahl,0.975)
datafile$zahl[datafile$zahl>upperQ] <- upperQ 

```


Standardize other headings
```{r}

datafile$MTB_Q <- datafile$quadrant
datafile$Fundort <-paste(datafile$fuo_gkk_rw,datafile$fuo_gkk_hw)
datafile$Species <- datafile$art
datafile$Beobachter <- datafile$projektgattung 

#change coordinates to long lat
#EPSG:31468
library(sp)
library(rgdal)
mycoords<- datafile[,c("fuo_gkk_rw","fuo_gkk_hw")]
coordinates(mycoords)<-c("fuo_gkk_rw","fuo_gkk_hw")
proj4string(mycoords)<-CRS("+init=epsg:31468")
mycoords <- spTransform(mycoords,CRS("+proj=longlat +datum=WGS84"))
datafile$lon <- mycoords@coords[,1]
datafile$lat <- mycoords@coords[,2]

#stage info here
unique(datafile$es)
#[1] "AD" "JU" "LK" "EX" "EI" "TA" "TL" "FR" "SC"

```

Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(Fundort))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))


library(GGally)
ggpairs(surveys[,5:9])


```

Characterize the total richness found in each grid:

```{r}


gridRichness <- ddply(datafile,.(MTB_Q,Month),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))


qplot(Month, nuSpecies,data=gridRichness)

gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 54

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

```


Get information on number of sampling locations on a given date

```{r}

samplingSites <- ddply(datafile,. (Date,MTB_Q),summarise,
nuLocations=length(unique(interaction(lon,lat))))

summary(samplingSites$nuLocations)
#between 1 and 30!!

#add to the data frame
datafile$samplingSites <- samplingSites$nuLocations[match(interaction(datafile$Date,datafile$MTB_Q),
                                interaction(samplingSites$Date,samplingSites$MTB_Q))]

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)


qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

#extract duplicates
library(gdata)

allObservers <- as.character(sort(unique(datafile$Beobachter)))

#for each observer, calculate how many records they were involved in collecting

expertise<-ldply(allObservers,function(x){
  Level = sum(grepl(x,datafile$Beobachter))
  data.frame(Beobachter=x,Level)
})
expertise<- subset(expertise,Level>0)

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

expertise$Level[expertise$Beobachter=="keine Zuordnung möglich"]<-1
expertise$Level[expertise$Beobachter=="sonstige Gutachten oder Auftragskartierungen"]<-median(expertise$Level)

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <-
  sapply(as.character(datafile$Beobachter),function(x){
    myobs <- as.character(x)
    max(expertise$Level[expertise$Beobachter %in% myobs])
    })

summary(datafile$Expertise)

```




