---
title: "preliminary_analysis_Odonata"
author: "Diana Bowler"
date: "24 januar 2018"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

#Analysis of the Odonata data from Saarland

First step: read in the raw data file from the sMon data portal.
Sheet 5 of the workbook contains the raw data of occurrences.
Each row in the data frame is a species observation.

```{r warning=FALSE}

library(gdata)
df <- read.xls("raw-data/Odonata_Saarland_Trockur.xls",sheet=5)

```

#Data frame formatting

Step one: format the date and extract month, dat and year information.

```{r}

library(lubridate)
df$Date <- as.Date(df$Date,file="%Y-%m-%d")
df$month <- month(df$Date)
df$day <- yday(df$Date)
df$year <- year(df$Date)

```

Step two: remove those with missing date entries

```{r}

df <- subset(df,!is.na(month))

```


#Data filtering

Filter one: focus on the months that were most sampled. The results suggest this is between April and September.

```{r}

table(df$month)

df <- subset(df, month %in% c(4:9))

```

Filter two: begin the time series from the first year with reasonable amount of data (i.e., number of records). 1981 seems like a good place to start. Also, we cut off 2017 incase the database is not up-to-date yet.

```{r}

table(df$Year)
df <- subset(df, Year>1980)
df <- subset(df, Year<2017)

```

#Data availability/Sampling effort checking

We can look at effort by: total number of records and average number of species seen per sampling day ("list length")

```{r}

library(sparta)

dataDiagnostics(taxa= as.character(df$Species), 
                site= as.character(df$MTB_Q), 
                time_period = df$Date)

```

Organise data using the sparta function

```{r}

tempDF <- formatOccData(taxa= as.character(df$Species), 
                site= as.character(df$MTB_Q), 
                time_period = df$Date,
                includeJDay = FALSE)


#get occurence matrix
occurrenceDF <- tempDF$spp_vis

#put out list length
listlengthDF <- tempDF$occDetdata

```


