---
title: "formatting_Saarland"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}

library(rgdal)
datafile <- read.csv("C:/Users/db40fysa/Nextcloud/sMon/sMon-Analyses/Odonata_Git/sMon-insects/raw-data/MV/qry_ArtData.csv",sep=";",dec=",",as.is=T)

source('C:/Users/db40fysa/Nextcloud/sMon/sMon-Analyses/Odonata_Git/sMon-insects/R/sparta_wrapper_functions.R')

```


Format data info

```{r}

library(lubridate)
datafile$Date <- sapply(datafile$Datum,function(x)strsplit(x," ")[[1]][1])
datafile$Date <- as.Date(datafile$Date,format="%d-%m-%Y")
datafile$Month <- month(datafile$Date)
datafile$Year <- year(datafile$Date)
datafile$yday <- yday(datafile$Date)

#and other columns
datafile$Species <- as.character(datafile$Art_Wissenschaftlich)
datafile$Beobachter <- datafile$GUID_Beobachter
datafile$Fundort <- datafile$Ortsbezeichnung
datafile <- subset(datafile, !is.na(yday))
datafile <- subset(datafile, yday!=1)

```

Sort MTBQ
```{r}

datafile <- subset(datafile,!is.na(MTB))
datafile <- subset(datafile,!is.na(MTBQ))
datafile$MTB_Q <- paste0(datafile$MTB,datafile$MTBQ)

```


Subset to just dragonflies

```{r}

#exclude: molch, kröte,schrecke,hüpfer (amphibians and grasshoppers)
notSpecies1 <- unique(datafile$Art_Deutsch[grepl("molch",datafile$Art_Deutsch)])
notSpecies2 <- unique(datafile$Art_Deutsch[grepl("kröte",datafile$Art_Deutsch)])
notSpecies3 <- unique(datafile$Art_Deutsch[grepl("schrecke",datafile$Art_Deutsch)])
notSpecies4 <- unique(datafile$Art_Deutsch[grepl("hüpfer",datafile$Art_Deutsch)])
notSpecies5 <- unique(datafile$Art_Deutsch[grepl("frosch",datafile$Art_Deutsch)])
notSpecies <- c(notSpecies1,notSpecies2,notSpecies3,notSpecies4,notSpecies5)

datafile <- subset(datafile, !Art_Deutsch %in% notSpecies)

```

Check species names against the standard list

```{r}
specieslist <- read.delim("specieslist_odonata.txt")

unique(datafile$Species[!datafile$Species%in%specieslist$Species])

#all good!
datafile <- subset(datafile, !Species %in% c("- Dummy-Art -","- Negativnachweis -"))

```

Subset to just adults
```{r}

einheitDF <- read.csv("C:/Users/db40fysa/Nextcloud/sMon/sMon-Analyses/Odonata_Git/sMon-insects/raw-data/MV/dbo_tblEinheiten.csv",sep=",",dec=",",as.is=T)
datafile$Einheit <- einheitDF$Einheit[match(datafile$ID_Einheit,einheitDF$ID_Einheit)]
table(datafile$Einheit)
unique(datafile$Einheit)
nrow(subset(datafile,!is.na(Einheit)))

datafile <- subset(datafile,!Einheit %in% c("Larve"))

```


Sort the anzahl column

```{r}

datafile$Anzahl_min <- datafile$Anzahl_Min
summary(datafile$Anzahl_min)
hist(datafile$Anzahl_min)
sum(datafile$Anzahl_min==0)
datafile <- subset(datafile,Anzahl_min!=0)#presence data

#bound extreme values
upperQ <- quantile(datafile$Anzahl_min,0.975,na.rm=T)
datafile$Anzahl_min[datafile$Anzahl_min>upperQ] <- upperQ 
summary(datafile$Anzahl_min)

```

Add lon/lat info

```{r}

datafile$lon <- datafile$xWGS84
datafile$lat <- datafile$yWGS84

```



Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(Fundort))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))


library(GGally)
ggpairs(surveys[,5:9])


```

Characterize the total richness found in each grid:

```{r}

gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 45

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

#add to data frame
datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

summary(datafile$Richness)


```


Get information on number of sampling locations on a given date

```{r}

samplingSites <- ddply(datafile,. (Date,MTB_Q),summarise,
nuLocations=length(unique(interaction(lon,lat))))

summary(samplingSites$nuLocations)
#1 to 22

```

Sort out the observer data:

```{r}

beobachterDF <- read.csv("C:/Users/db40fysa/Nextcloud/sMon/sMon-Analyses/Odonata_Git/sMon-insects/raw-data/MV/dbo_vwBeobachter.csv",sep=",",as.is=T)

datafile$Beobachter <- beobachterDF$ID_Beobachter[match(datafile$GUID_Beobachter,beobachterDF$GUID_Beobachter)]
#211 is unknown
datafile$Beobachter[is.na(datafile$Beobachter)] <- 211

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

allObservers <- as.character(sort(unique(datafile$Beobachter)))


#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise<-ldply(allObservers,function(x){
  Beobachter <- as.character(datafile$Beobachter)
  Beobachter <- removeGerman(Beobachter)
  Level = sum(x==Beobachter)
  data.frame(Beobachter=x,Level)
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

#give 1 to the unknown observers
expertise$Beobachter <- as.character(expertise$Beobachter)
expertise$Beobachter[expertise$Beobachter==211]<-1

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <- expertise$Level[match(datafile$Beobachter,expertise$Beobachter)]
datafile$Expertise[is.na(datafile$Expertise)]<-1
datafile$Expertise[is.infinite(datafile$Expertise)]<-1

summary(datafile$Expertise)


```


Save data

```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","Anzahl_min")]
save(adult_datafile,file="derived-data/adult_datafile_MV.RData")
saveRDS(adult_datafile,file="derived-data/adult_datafile_MV.rds")

```



