---
title: "formatting_BW"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in original data frame

```{r}

datafile <- read.csv("raw-data/BW/SGL_Libellen_2006_bis_2018.csv",as.is=T,sep=";")
nrow(datafile)
#42747

source('C:/Users/db40fysa/Nextcloud/sMon/sMon-Analyses/Odonata_Git/sMon-insects/R/sparta_wrapper_functions.R')

```

plot these coordinates just for a reference

```{r}

library(sp)

#"SGL-Schutzgemeinschaft Libellen BW"

unique(datafile$EPSG)#31467
coords <- datafile[,c("X","Y")]
coordinates(coords) <- c("X","Y")
proj4string(coords) <- CRS("+init=epsg:31467")
plot(coords)

#get shapefile for Baden Wuttemburg

germanAdmin <- readRDS("C:/Users/db40fysa/Nextcloud/sMon/sMon-Analyses/Spatial_data/AdminBoundaries/gadm36_DEU_1_sp.rds")
BW <- subset(germanAdmin,NAME_1 %in% c("Baden-Württemberg"))
BW <- spTransform(BW,CRS("+init=epsg:31467"))

plot(BW)
plot(coords,add=T)

```


get new data file

```{r}

datafile <- read.csv("raw-data/update 2020/BW/SGL_Daten_bis 2005.csv",as.is=T,sep=",")
nrow(datafile)#83928

```


Format data info

```{r}

#any missing data
sum(is.na(datafile$Datum))
sum((datafile$Datum==""))#24603

library(lubridate)
datafile <- subset(datafile, Datum!="")
datafile$Date <- sapply(datafile$Datum,function(x)strsplit(x," ")[[1]][1])
datafile$Date <- as.Date(datafile$Date,format="%d.%m.%Y")
datafile$Month <- month(datafile$Date)
datafile$Year <- year(datafile$Date)
datafile$yday <- yday(datafile$Date)
datafile <- subset(datafile,yday!=1)

nrow(datafile)#59309

```

Fix column headings
```{r}

datafile$Species <- datafile$Art.Wissenschaftlich

```

Fix MTBQ
```{r}

datafile$MTB_Q <- apply(datafile,1,function(x)paste0(x["MTB"],x["MTBQ"]))

```

Check species names against the standard list

```{r}

specieslist <- read.delim("specieslist_odonata.txt")
unique(datafile$Species[!datafile$Species%in%specieslist$Species])


```

Identifying adult observation

```{r}

unique(datafile$Einheit)
table(datafile$Einheit)

datafile <- subset(datafile, !Einheit %in% c("Larve"))

```

Sort count data

```{r}

#check for extremes
hist(datafile$Anzahl)
summary(datafile$Anzahl)

#block upper values
datafile$Anzahl[datafile$Anzahl > quantile(datafile$Anzahl,0.99,na.rm=T)] <- quantile(datafile$Anzahl,0.99,na.rm=T)
summary(datafile$Anzahl)

```

Get coords of the MTBs

```{r}

load("mtbqsDF.RData")
datafile$lon <- mtbqsDF$lon[match(datafile$MTB_Q,mtbqsDF$MTB_Q)]
datafile$lat <- mtbqsDF$lat[match(datafile$MTB_Q,mtbqsDF$MTB_Q)]

```


Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(lat,lon))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))

library(GGally)
ggpairs(surveys[,5:9])

```

Characterize the total richness found in each grid:

```{r}
gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 46
#1 is unlikely...

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

#add to data frame
datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

summary(datafile$Richness)

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

allObservers <- as.character(sort(unique(datafile$Beobachter)))

#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise <-ldply(allObservers,function(x){
  data.frame(Beobachter=x,Level=sum(x==datafile$Beobachter))
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <- expertise$Level[match(datafile$Beobachter,expertise$Beobachter)]

summary(datafile$Expertise)

```

#subset to columns of interest

```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","Anzahl")]
names(adult_datafile)[which(names(adult_datafile)=="Anzahl")] <- "Anzahl_min"


```

Get data from the previous year

```{r}

adult_datafile_orig <- readRDS("derived-data/adult_datafile_BW.rds")

#remove duplicates
adult_datafile <- rbind(adult_datafile,adult_datafile_orig) 
adult_datafile <- unique(adult_datafile)

```

Save juvenile and adult data separately

```{r}

saveRDS(adult_datafile,file="derived-data/adult_datafile_BW_updated.rds")

```

plot data

```{r}

library(sp)

germanAdmin <- readRDS("C:/Users/db40fysa/Nextcloud/sMon/sMon-Analyses/Spatial_data/AdminBoundaries/gadm36_DEU_1_sp.rds")
BW <- subset(germanAdmin,NAME_1 %in% c("Baden-Württemberg"))

coords <- adult_datafile[,c("lon","lat")]
coordinates(coords) <- c("lon","lat")
proj4string(coords) <- CRS(proj4string(BW))

#plot it
plot(BW)
plot(coords,add=T)
#looks good!!!

```


