---
title: "formatting_Saarland"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}

library(readxl)

datafile <- read_excel("raw-data/NiederSaschen/000_NDS_HB_data_export_iDiv_w_date.xlsx")

source('C:/Users/db40fysa/Nextcloud/sMon-Analyses/Git/sMon-insects/R/sparta_wrapper_functions.R')

```

Format data info

```{r}

library(lubridate)
datafile$Date <- as.Date(datafile$date,format="%Y-%m-%d")
datafile$Month <- month(datafile$Date)
datafile$Year <- year(datafile$Date)
datafile$yday <- yday(datafile$Date)

```

Fix column headings
```{r}

datafile$Species <- datafile$species
datafile$MTB_Q <- datafile$mtbq
datafile$Beobachter <- datafile$observer_id

#save natural region info
naturraum_Saarland <- unique(datafile[,c("MTB_Q","nat_reg_id")])
saveRDS(naturraum_Saarland,file="naturraum_Saarland.rds")

```


Check species names against the standard list

```{r}
specieslist <- read.delim("specieslist_odonata.txt")


unique(datafile$Species[!datafile$Species%in%specieslist$Species])

#correct species names
datafile$Species[which(datafile$Species=="Lestes virens vestalis")]<-"Lestes virens"
datafile$Species[which(datafile$Species=="Aeshna subarctica elisabethae")]<-"Aeshna subarctica"

unique(datafile$Species[!datafile$Species%in%specieslist$Species])

```

Identifying adult observation

```{r}

unique(datafile$unit)
table(datafile$unit)

datafile <- subset(datafile, !unit %in% c("Larve","Jungtier"))

```

Restrict to adult sampling

```{r}
#check for extremes
hist(datafile$count)
summary(datafile$count)
#quantile(datafile$count,0.99)

#remove zero data
nrow(subset(datafile,count==0))
datafile <- subset(datafile,count!=0)

#block upper values
datafile$count[datafile$count > quantile(datafile$count,0.99,na.rm=T)] <- quantile(datafile$count,0.99,na.rm=T)
hist(datafile$count)

```

Get coords of the MTBs

```{r}
load("mtbqsDF.RData")
datafile$lon <- mtbqsDF$x[match(datafile$MTB_Q,mtbqsDF$MTB_Q)]
datafile$lat <- mtbqsDF$y[match(datafile$MTB_Q,mtbqsDF$MTB_Q)]
datafile <- subset(datafile,!is.na(lon))

```


Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(lat,lon))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))

library(GGally)
ggpairs(surveys[,5:9])

```

Characterize the total richness found in each grid:

```{r}
gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 51!!!
#1 is unlikely...

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

#add to data frame
datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

summary(datafile$Richness)

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)
#1 to 66!!!

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

allObservers <- as.character(sort(unique(datafile$Beobachter)))

#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise <-ldply(allObservers,function(x){
  data.frame(Beobachter=x,Level=sum(x==datafile$Beobachter))
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <- expertise$Level[match(datafile$Beobachter,expertise$Beobachter)]

summary(datafile$Expertise)

```

Save juvenile and adult data separately

```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","count")]
names(adult_datafile)[which(names(adult_datafile)=="count")] <- "Anzahl_min"
save(adult_datafile,file="derived-data/adult_datafile_NS.RData")
saveRDS(adult_datafile,file="derived-data/adult_datafile_NS.rds")

```



