---
title: "formatting_Saarland"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}

datafile <- read.delim("raw-data/He/X_Jena.txt",as.is=T)
datafile2 <- read.delim("raw-data/He/X_Jena2.txt",as.is=T)
datafile <- unique(rbind(datafile,datafile2))

source('C:/Users/db40fysa/Nextcloud/sMon-Analyses/Git/sMon-insects/R/sparta_wrapper_functions.R')

```

Format data info

```{r}

library(lubridate)
datafile$Date <- sapply(datafile$Datum,function(x)strsplit(x," ")[[1]][1])
datafile$Date <- as.Date(datafile$Date,format="%m/%d/%Y")
datafile$Month <- month(datafile$Date)
datafile$Year <- year(datafile$Date)
datafile$yday <- yday(datafile$Date)

```

Fix column headings
```{r}

datafile$Species <- datafile$Art

```

Fix MTBQ
```{r}

datafile$Q <- sapply(as.character(datafile$Rasterfeld),function(x)substr(x,1,1))
datafile$MTB_Q <- apply(datafile,1,function(x)paste0(x["MTB"],x["Q"]))

```

Check species names against the standard list

```{r}
specieslist <- read.delim("specieslist_odonata.txt")
unique(datafile$Species[!datafile$Species%in%specieslist$Species])

#remove genus data
datafile <- datafile[grepl(" ",datafile$Species),]
  
#correct species names
datafile$Species[which(datafile$Species=="Pyrrhosoma nympula")]<-"Pyrrhosoma nymphula"

#check
unique(datafile$Species[!datafile$Species%in%specieslist$Species])

```

Identifying adult observation

```{r}

unique(datafile$Einheit)
table(datafile$Einheit)

datafile <- subset(datafile, Einheit %in% c("Alttier/Imago","Emergenz","Eiablage",
                                            "Keine Angabe","Exuvien"))

```

Sort count data

```{r}


datafile$Anzahl_min <- NA

```

Get coords of the MTBs

```{r}
load("mtbqsDF.RData")
datafile$lon <- mtbqsDF$x[match(datafile$MTB_Q,mtbqsDF$MTB_Q)]
datafile$lat <- mtbqsDF$y[match(datafile$MTB_Q,mtbqsDF$MTB_Q)]
datafile <- subset(datafile,!is.na(lon))

```


Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(lat,lon))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))

library(GGally)
ggpairs(surveys[,5:9])

```

Characterize the total richness found in each grid:

```{r}
gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 50
#1 is unlikely...

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

#add to data frame
datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

summary(datafile$Richness)

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)
#1 to 60!!!

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

allObservers <- as.character(sort(unique(datafile$Beobachter)))

#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise <-ldply(allObservers,function(x){
  data.frame(Beobachter=x,Level=sum(x==datafile$Beobachter))
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <- expertise$Level[match(datafile$Beobachter,expertise$Beobachter)]

summary(datafile$Expertise)

```

Save juvenile and adult data separately

```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","Anzahl_min")]
save(adult_datafile,file="derived-data/adult_datafile_He.RData")
saveRDS(adult_datafile,file="derived-data/adult_datafile_He.rds")

```



