---
title: "formatting_Saarland"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}

#read in first datafile sent
datafile <- read.delim("raw-data/update 2020/He/Daten_1980-2019.csv",as.is=T,sep=",")
datafile <- datafile[,-which(names(datafile)=="R16")]

source('C:/Users/db40fysa/Nextcloud/sMon/sMon-Analyses/Odonata_Git/sMon-insects/R/sparta_wrapper_functions.R')

```

read in also additional file

```{r}

#read in second data file sent with higher resolution data
datafile2 <- read.delim("raw-data/update 2020/He/Daten_2006-2019.txt",sep=",",as.is=T)
datafile2$Datum <- sapply(datafile2$Datum,function(x)strsplit(x," ")[[1]][1])

#subset to same columns in the first file
datafile2 <- datafile2[,names(datafile)]

#and merge
datafile <- rbind(datafile,datafile2)

#remove duplicates
datafile <- unique(datafile)

```


Format data info

```{r}

library(lubridate)
datafile$Date <- as.Date(datafile$Datum,format="%m/%d/%Y")
datafile$Month <- month(datafile$Date)
datafile$Year <- year(datafile$Date)
datafile$yday <- yday(datafile$Date)
datafile <- subset(datafile,yday!=1)
datafile <- subset(datafile,!is.na(yday))

```

Fix column headings
```{r}

datafile$Species <- datafile$Art

```

Fix MTBQ
```{r}

datafile$MTB_Q <- apply(datafile,1,function(x)paste0(x["MTB"],x["Q"]))

```

Check species names against the standard list

```{r}
specieslist <- read.delim("specieslist_odonata.txt")
unique(datafile$Species[!datafile$Species%in%specieslist$Species])

#remove higher-level data
datafile <- subset(datafile,!Species %in% c("Odonata","Cordulegaster","Calopteryx","Sympetrum","Aeshna","Erythromma"))
  
```

Identifying adult observation

```{r}

unique(datafile$Einheit)
table(datafile$Einheit)

datafile <- subset(datafile, !Einheit %in% c("Larve"))

```

Sort count data

```{r}

datafile$Anzahl_min <- datafile$Anzahl
summary(datafile$Anzahl_min)

```

Get coords of the MTBs

```{r}

load("mtbqsDF.RData")
datafile$lon <- mtbqsDF$x[match(datafile$MTB_Q,mtbqsDF$MTB_Q)]
datafile$lat <- mtbqsDF$y[match(datafile$MTB_Q,mtbqsDF$MTB_Q)]

```


Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(lat,lon))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))

library(GGally)
ggpairs(surveys[,5:9])

```

Characterize the total richness found in each grid:

```{r}

gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 50
#1 is unlikely...

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

#add to data frame
datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

summary(datafile$Richness)

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)
#1 to 60!!!

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

allObservers <- as.character(sort(unique(datafile$Beobachter)))

#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise <-ldply(allObservers,function(x){
  data.frame(Beobachter=x,Level=sum(x==datafile$Beobachter))
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <- expertise$Level[match(datafile$Beobachter,expertise$Beobachter)]

summary(datafile$Expertise)

```

limit to standard columns
```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","Anzahl_min")]

```


combine with previous data for Hessen

```{r}

adult_datafile_orig <- readRDS("derived-data/adult_datafile_He.rds")

#remove duplicates
adult_datafile <- rbind(adult_datafile,adult_datafile_orig) 
adult_datafile <- unique(adult_datafile)

```


```{r}

#save(adult_datafile,file="derived-data/adult_datafile_He_updated.RData")
saveRDS(adult_datafile,file="derived-data/adult_datafile_He_updated.rds")

```

#Jorg says
#For our Red List we are mainly looking at the period from 2006 onwards, 
#so it would be great if you could do the analysis for this #period and in MTB/64 raster
#so we also save:

```{r}

datafile_64 <- read.delim("raw-data/update 2020/He/Daten_2006-2019.txt",sep=",",as.is=T)

#format Date
datafile_64$Datum <- sapply(datafile_64$Datum,function(x)strsplit(x," ")[[1]][1])
datafile_64$Date <- as.Date(datafile_64$Datum,format="%m/%d/%Y")
datafile_64$Month <- month(datafile_64$Date)
datafile_64$Year <- year(datafile_64$Date)
datafile_64$yday <- yday(datafile_64$Date)

#species
datafile_64$Species <- datafile_64$Art
unique(datafile_64$Species[!datafile_64$Species%in%specieslist$Species])
datafile_64 <- subset(datafile_64,!Species %in% c("Odonata","Cordulegaster","Calopteryx","Sympetrum","Aeshna","Erythromma"))

#unit
datafile_64 <- subset(datafile_64, Einheit %in% c("Alttier/Imago","Emergenz","Keine Angabe","Exuvien"))
datafile_64$Anzahl_min <- datafile_64$Anzahl

#select subset of variables
datafile_64$MTB_Q <- apply(datafile_64,1,function(x)paste0(x["MTB"],x["Q"]))

adult_datafile_64<- datafile_64[,c("Date","MTB","MTB_Q","MTB64","Species","Beobachter","Anzahl_min")]


saveRDS(adult_datafile_64,file="derived-data/adult_datafile_He_updated_MTB64.rds")
