---
title: "formatting_Bavaria"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Read in data frame

```{r}

library(rgdal)

datafile <- readOGR(dsn="C:/Users/db40fysa/Nextcloud/sMon/sMon-Analyses/Odonata_Git/sMon-insects/raw-data/Saxony/Libellen",
                    layer="20181122-ZeNA-Libellen-3")

#convert to data frame
datafile <- datafile@data

```

Add species names

```{r}

datafile$Species <- datafile$Art_wiss

```

Check species names against the standard list

```{r}
specieslist <- read.delim("model-auxfiles/specieslist_odonata.txt")

unique(datafile$Species[!datafile$Species%in%specieslist$Species])

#remover genus only
datafile <- subset(datafile,grepl(" ",Species))
unique(datafile$Species[!datafile$Species%in%specieslist$Species])

```

Check out geographic coords

```{r}

sort(unique(datafile$Toleranz))
table(datafile$Toleranz)

datafile <- subset(datafile, Toleranz!="Raster MTB")

```

Add MTB_Q info

```{r}
library(rgdal)

#get MTB Q
datafile$MTB_Q <- as.numeric(paste0(as.character(datafile$MTB),as.character(datafile$MTBQ)))

#add lat lon coords
mycoords <- datafile[,c("X","Y")]
coordinates(mycoords) <-c("X","Y")
#proj4string(mycoords)<-CRS("+proj=tmerc +lat_0=0 +lon_0=6 +k=1 +x_0=2500000 +y_0=0 +ellps=bessel +datum=potsdam +units=m +no_defs")
proj4string(mycoords)<-CRS("+proj=utm +zone=33 +ellps=GRS80 +units=m +no_defs")
mycoords <- spTransform(mycoords,CRS("+proj=longlat +datum=WGS84"))
head(mycoords)
datafile$lon <- mycoords@coords[,1]
datafile$lat <- mycoords@coords[,2]
datafile <- subset(datafile, !is.na(MTB_Q))

```

Sort out the dates

```{r}

library(lubridate)
datafile$Date <- sapply(as.character(datafile$Datum),function(x)strsplit(x," ")[[1]][1])
datafile$Date <- as.Date(datafile$Date,format="%d.%m.%Y")
datafile$Month <- month(datafile$Date)
datafile$Year <- year(datafile$Date)
datafile$yday <- yday(datafile$Date)
summary(datafile$yday)

nrow(subset(datafile,is.na(yday)))
nrow(subset(datafile,!is.na(yday)))

datafile <- subset(datafile, !is.na(yday))
datafile <- subset(datafile, yday!=1)

```

Standardize other headings

```{r}

datafile$Fundort <-paste(datafile$X,datafile$Y)
datafile$Beobachter <- datafile$MTB

```


Sort stage info and restrict to adult

```{r}

unique(datafile$Einheit)
table(datafile$Einheit)

datafile <- subset(datafile,!Einheit %in%c("Larve"))

```


Sort data columns

```{r}

hist(datafile$Anzahl)
summary(datafile$Anzahl)

upperQ <- quantile(datafile$Anzahl,0.975,na.rm=T)
datafile$Anzahl[datafile$Anzahl>upperQ] <- upperQ 
summary(datafile$Anzahl)

```

Get survey information

```{r}

library(plyr)

surveys <- ddply(datafile,.(MTB_Q,Date,Year,Month,yday),summarise,
                 nuLocations = log(length(unique(Fundort))),
                 nuObservers = log(length(unique(Beobachter))),
                 nuSpecies = log(length(unique(Species))),
                 nuRecords = log(length(Species)))


library(GGally)
ggpairs(surveys[,5:9])


```

Characterize the total richness found in each grid:

```{r}


gridRichness <- ddply(datafile,.(MTB_Q,Month),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))


qplot(Month, nuSpecies,data=gridRichness)

gridRichness <- ddply(datafile,.(MTB_Q),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species),
                    nuYears = length(unique(Year)),
                    nuDates = length(unique(Date)))

summary(gridRichness$nuSpecies)
#1 to 48

qplot(log(nuRecs),log(nuSpecies),data=gridRichness)
#the more records, the more species

datafile$Richness <- gridRichness$nuSpecies[match(datafile$MTB_Q,gridRichness$MTB_Q)]

```


Get information on number of sampling locations on a given date

```{r}

samplingSites <- ddply(datafile,. (Date,MTB_Q),summarise,
nuLocations=length(unique(interaction(lon,lat))))

summary(samplingSites$nuLocations)
#between 1 and 88

#add to the data frame
datafile$samplingSites <- samplingSites$nuLocations[match(interaction(datafile$Date,datafile$MTB_Q),
                                interaction(samplingSites$Date,samplingSites$MTB_Q))]

```

Sort out the observer data:

```{r}

sort(unique(datafile$Beobachter))

```

Characterize the total ability of each observer each year

```{r}

obsRichness <- ddply(datafile,.(Beobachter),summarise,
                    nuSpecies=length(unique(Species)),
                    nuRecs=length(Species))

summary(obsRichness$nuSpecies)
#1 to 70

qplot(log(nuRecs),log(nuSpecies),data=obsRichness)

arrange(obsRichness,desc(nuRecs))

```

Characterise average expertise level

```{r}

allObservers <- as.character(sort(unique(datafile$Beobachter)))

#for each observer, calculate how many records they were involved in collecting
library(plyr)
expertise <-ldply(allObservers,function(x){
  data.frame(Beobachter=x,Level=sum(x==datafile$Beobachter))
})

hist(log(expertise$Level))  

arrange(expertise,desc(Level))

```

Add information on expertise to the datafile

```{r}

datafile$Expertise <- expertise$Level[match(datafile$Beobachter,expertise$Beobachter)]

summary(datafile$Expertise)

```


Save data

```{r}

adult_datafile<- datafile[,c("Date","MTB_Q","Species","Expertise",
                           "Richness","lon","lat","Beobachter","Anzahl")]
names(adult_datafile)[which(names(adult_datafile)=="Anzahl")] <- "Anzahl_min"
save(adult_datafile,file="derived-data/adult_datafile_Sax.RData")
saveRDS(adult_datafile,file="derived-data/adult_datafile_Sax.rds")

```

