---
title: "preliminary_analysis_Odonata"
author: "Diana Bowler"
date: "24 januar 2018"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

#Analysis of the Odonata data 

First step: read in the raw data file from the sMon data portal.
Sheet 5 of the workbook contains the raw data of occurrences.
Each row in the data frame is a species observation.

Select state for analysis

```{r}

state = "Sa"
state = "SH"
state = "Bav"
state = "NRW"
state = "Sax"
state="SAnhalt"
state="RLP"
state="Thu"
state="BB"
state="NS"
state="MV"
state="He"
state="BW"
```

Read in the dataset

```{r warning=FALSE}

load(paste0("derived-data/adult_datafile_",state,".RData"))
df <- adult_datafile

```


#Data frame formatting

Step one: format the date and extract month, dat and year information.

```{r}

library(lubridate)
df$Date <- as.Date(df$Date,file="%Y-%m-%d")
df$month <- month(df$Date)
df$day <- yday(df$Date)
df$Year <- year(df$Date)

#remove records on 1st Jan - prob inaccurate...
nrow(subset(df,(month==1&day==1)))
df <- subset(df,!(month==1&day==1))

```

Step two: remove those with missing date entries

```{r}

df <- subset(df,!is.na(day))
df <- subset(df,!is.na(Year))
df <- subset(df,!is.na(month))
summary(df)

```

Look at summary info per species

```{r}

library(plyr)
outSummary <- ddply(df,.(Species),summarise,
                     nuRecs=length(unique(Date)),
                     nuYears=length(unique(Year)),
                     nuSites=length(unique(MTB_Q)))

arrange(outSummary,nuYears)

```

#Data filtering

Filter by year

Begin the time series from the first year with reasonable amount of data (i.e., number of records). 1981 seems like a good place to start. Also, we cut off 2017 incase the database is not up-to-date yet.

Start from first year with 50 records 
```{r}

table(df$Year)
df <- subset(df,Year>1979 & Year <2018)
surveyYears <- sort(unique(df$Year))
#start year is first year 1980 on with >50 records
possibleYears <- names(table(df$Year))[table(df$Year)>50]
firstYear <- min(possibleYears)
otherYears <- surveyYears[surveyYears>firstYear]
yearDF <- data.frame(Year=as.numeric(as.character(c(firstYear,otherYears))),
                     State=state)
write.table(yearDF,file=paste0("yearDF_","adult","_",state,".txt"),
            sep="\t",row.names=FALSE)

```

Remove rarely visited sites?? no

```{r}

sitesSummary<-ddply(df,.(MTB_Q),summarise,nuYears=length(unique(Year)))
#df <- subset(df, MTB_Q %in% siteSummary)

```

How much data is there across large scale grids
```{r}

load("mtbqsDF.RData")

```


#Filter by flight period for each species

Fits a gam for each species

```{r}

#insert 0s for when there was a survey

allSurveys <- expand.grid(Date=unique(df$Date),Species=unique(df$Species))
df0 <- merge(allSurveys,df,by=c("Species","Date"),all.x=T)
df0$Anzahl_min[interaction(df0$Date,df0$Species) %in% interaction(df$Date,df$Species)]<-1
df0$Anzahl_min[!interaction(df0$Date,df0$Species) %in% interaction(df$Date,df$Species)]<-0
df0$month <- month(df0$Date)
df0$day <- yday(df0$Date)
df0$Year <- year(df0$Date)

#look at plots
#library(ggplot2)
#ggplot(df0,aes(x=day,y=Anzahl_min))+
#  stat_smooth(method="loess")+
#  facet_wrap(~Species,scales="free")


#ggplot(df0,aes(x=day,y=Anzahl_min))+
#  stat_smooth(formula = y ~ poly(x,2),method="glm",method.args = list(family = #"binomial"))+
#  facet_wrap(~Species,scales="free")

```


Fit a model with cyclic cubic regression splines

```{r}
library(mgcv)
df0$fYear <- as.factor(df0$Year)
df0 <- subset(df0, Species %in% outSummary$Species[outSummary$nuRecs>25])
#species 1 not fit for SH
#species 9 doesnt work for Bav
#species 46 doesnt work for Sax
#species 53 doesnt for SAnhalt
#species 58 doesnt work for BB
out <- ddply(df0,.(Species),function(x){
             gam1 <- gamm(Anzahl_min ~  1 + s(day, bs="cc"), 
                          random = list(fYear=~1), 
                          family=binomial, data=x)
             fits <- data.frame(day=min(x$day):max(x$day),                              fits=predict(gam1$gam,newdata=data.frame(day=c(min(x$day):max(x$day))),type="response"))
             return(fits)
            })

#ggplot(out)+
#  geom_line(aes(x=day,y=fits))+
#  facet_wrap(~Species,scales="free")

```

Identify days when species occurence is above 5%

```{r}
speciesDays <- subset(out,fits>0.01)

ggplot(speciesDays)+
  geom_point(aes(x=day,y=fits))+
  facet_wrap(~Species,scales="free")

write.table(speciesDays,file=paste0("speciesDays_adult_",state,".txt"),
            sep="\t",row.names=FALSE)

```

Look at months when people sample

```{r}

monthSummary <- ddply(df,.(month),summarise,
                      nuBeobachter=length(unique(Beobachter)),
                      nuYears=length(unique(Year)),
                      nuSpecies=length(unique(Species)),
                      nuVisits=length(Species),
                      nuMTBQ=length(unique(MTB_Q)))
monthSummary

#subset to months
monthSummary <- subset(monthSummary, nuYears > (length(unique(df$Year))*0.5) &
                         nuMTBQ > (max(monthSummary$nuMTBQ)*0.5))
monthDF <- data.frame(startMonth = min(monthSummary$month),
                      endMonth = max(monthSummary$month),
                      State=state)
monthDF

write.table(monthDF,file=paste0("monthDF_","adult_",state,".txt"),
            sep="\t",row.names=FALSE)

```


More Summary info per species

```{r}

library(plyr)
(outSummary <- ddply(df,.(Species),summarise,
                     nuRecs=length(unique(Date)),
                     nuYears=length(unique(Year)),
                     nuSites=length(unique(MTB_Q)),
                      nuRepeats=length(Anzahl_min)/length(unique(Year))))#repeat obs within a year

#get species seen in 25% of years and species usually seen at least twice
mainSpecies <- outSummary$Species[(outSummary$nuYears > (length(unique(df$Year))/4)) & (outSummary$nuRepeats>2)]

#also those with sufficient data to estimate their phenology
mainSpecies <- mainSpecies[mainSpecies%in%speciesDays$Species]

mainSpeciesDF <- data.frame(TaskID=1:length(mainSpecies),Species=mainSpecies)
mainSpecies
write.table(mainSpeciesDF,file=paste0("speciesTaskID_adult_",state,".txt"),sep="\t",row.names=FALSE)

```

Organise data using the sparta wrapper function

```{r, results='asis'}

source('R/sparta_wrapper_functions.R')

#define a visit
#df$visit <- paste(df$MTB_Q,df$Date,df$Beobachter,sep="_")
df$visit <- paste(df$MTB_Q,df$Date,sep="_")

occMatrix <- getOccurrenceMatrix(df)
listlengthDF <- getListLength(df)

#checks
all(row.names(occMatrix)==listlengthDF$surveyList)
listlengthDF[1:10,]

#plots
library(ggplot2)
qplot(nuSpecies,nuRecords,data=listlengthDF)
qplot(nuSpecies,RpS,data=listlengthDF)

```


#Run models over multiple species

Pull out 20 most common Species

```{r}

outSummary <- arrange(outSummary,desc(nuRecs))
(mySpecies <- outSummary$Species[1:20])
outSummary$nuRecs[1:20]

```

Run model with effort with all effort terms

```{r}

#run out each species
out <-llply(mySpecies,function(x){
  fitBugs(mySpecies=x)
})

save(out,file=paste0("out_nuSpecies_All_adults_",state,".RData"))

```


Run model with number of records as effort term

```{r}

#run out each species
out <-llply(mainSpecies,function(x){
  fitBugs(mySpecies=x,effort="nuRecs")
})

save(out,file=paste0("out_nuRecords_adults_",state,".RData"))

```

Individuals fits
```{r}
out <- fitBugs(mySpecies="Cordulia aenea")

```

Get models and save annual indicies

```{r}
#get annual indicies
load(paste0("model-outputs/out_nuRecords_adults_",state,".RData"))
out2 <- combineModels(out)
outTrends <- getBUGSfits(out2,param="psi.fs") 

save(outTrends,file=paste0("outTrends_nuRecords_adult_",state,".RData"))

```

Test effect of fitting a spatial model
```{r}

#fit a spatial spline to occurrence probability
listlengthDF$obs <- occMatrix[,"Aeshna cyanea"]

#get x and y coords
listlengthDF$MTB_Q <- gsub("/","",listlengthDF$MTB_Q)
load("mtbqsDF.RData")
listlengthDF$x <- mtbqsDF$x[match(listlengthDF$MTB_Q,mtbqsDF$MTB_Q)]
listlengthDF$y <- mtbqsDF$y[match(listlengthDF$MTB_Q,mtbqsDF$MTB_Q)]

library(mgcv)
out1<-gam(obs ~ 1 + s(x,y),data=listlengthDF,family=binomial)
listlengthDF$fits <- out1$fitted.values

library(ggplot2)
qplot(x,y,data=listlengthDF,colour=fits)+
  scale_colour_gradient(low="blue",high="red")

```

Model name meaning on HPC
```{r}
#out_
#restrict to common months

#out_flightperiod
#use any data falling within the species flight period

#our_dynamic
#separate colonization and persistence

```
